---
title: "Analyzing dictionary with recursive functions in R"
author: "Pritesh Shrivastava"
date: "12/5/2020"
output: md_document
---

The following exercise problems are from the introductory Computer Science book, How to Design Programs. The problems were originally defined in a custom langauge built from the Racket programming language. These problems employed functional programming & recursion to build some analytic fucntions from the scratch & I really enjoyed working on them. In this post, I'll solve these problems in R instead.


### Reading from the computer's dictionary

You don't need any external data to work on these problems. Your computer's built-in dictionary is good enough. Depending on your OS, you can find your dictionary in one of the following locations.
```{r}
# On OS X: 
LOCATION = "/usr/share/dict/words"
# On LINUX: /usr/share/dict/words or /var/lib/dict/words
# On WINDOWS: borrow the word file from your Linux friend

dictionary_list <- scan(LOCATION, what = "character", sep="\n")
dictionary_list[0:5]
```

```{r}
length(dictionary_list)
```

But our list is much longer than 3000! So we can't use normal recursive functions on this list, otherwise, it'll cause a [stack overflow](https://stackoverflow.com/questions/14222416/recursion-in-python-runtimeerror-maximum-recursion-depth-exceeded-while-callin) !!
We can solve this problem by either writing iterative procedures instead, or converting our recursive procuderes to their tail-call form. We'll also be adding decorators to our function to ensure Python optimizes our tail-recursive calls.

### Problem 1

Design the function `starts_with`, which consumes a Letter and Dictionary and then counts how many words in the given Dictionary start with the given Letter.
```{r}
trampoline <- function(f, ...) {
  function(...) {
    ret <- f(...)
    while (inherits(ret, "recursion")) {
      ret <- eval(as.call(c(f, unclass(ret))))
    }
    ret
  }
}

recur <- function(...) {
  structure(list(...), class = "recursion")
}
## Source : https://tailrecursion.com/wondr/posts/tail-recursion-in-r.html
```

```{r}
starts_with <- trampoline(function(letter, dict_list, accumulator=0) {
    ## Counts how many words in the dict_list start with the letter let
    if(length(dict_list) == 0) accumulator
    else{
        if(dict_list[1][1] == letter){ ## Indices in R start from 1, not 0
            recur(letter, tail(dict_list, -1), accumulator+1)
        }else{
            recur(letter, tail(dict_list, -1), accumulator)
          }
        }
})

## Testing function on a couple of letter values
starts_with('a', dictionary_list)
starts_with('z', dictionary_list)
```

